\begin{abstract}

% Reduce motivation?
\textit{Live programming environments} provide various semantic services, including type checking and evaluation, continuously as the user is editing the program. The live paradigm promises to improve the  developer experience, but liveness is an implementation challenge, particularly when working with large programs. This paper specifies and efficiently implements a system that is able to incrementally update type information for a live program in response to fine-grained program edits. This information includes type error marks and information about the expected and actual type of every expression. The system is specified type-theoretically as a small-step dynamics that propagates updates through the marked and annotated program. Most updates flow according to a base bidirectional type system. Additional pointers are maintained to connect bound variables to their binding locations, with type updates traversing these pointers directly. Order maintenance data structures are employed to efficiently maintain these pointers and to prioritize the order of update propagation. We prove this system is equivalent to naive reanalysis in the Agda theorem prover, along with other important metatheoretic properties. We then provide an efficient OCaml implementation, detailing a number of impactful optimizations. We evaluate this implementation's performance with a large stress-test and find that it is able to achieve multiple orders of magnitude speed-up compared to from-scratch reanalysis. 

\end{abstract}